# About {.unnumbered}

The field of thermoregulation is fundamental to photosynthesis but the lack of understanding and process-based modelling of leaf temperature regulation creates a knowledge gap worthy of investigation. Moreover, by exploiting EEO principles, we can create a model that accounts for the interplay of thermal acclimation and thermoregulation. Understanding this interplay is an unresolved but highly relevant question because it gives insights into the processes a plant may use to cope with a rising temperatures [@cavaleri2020cold].

This tutorial provides the scientific basis and code implementation to model a leaf's temperature based on thermodynamic theory and eco-evolutionary optimality theory. @sec-motivation gives insight into the current debate on leaf thermoregulation and makes the statement how eco-evolutionary optimality theory may provide a solution to this issue. In @sec-algorithm, a numerical algorithm is presented alongside a sensitivity analysis showing the strengths and weaknesses of the current implemenation. In @sec-next-steps, further improvements of the algorithm and ideas for potential test-cases to evaluate the model are provided.

# Requirements

```{r include=TRUE}
source(here::here("R/_setup.R"))
```

```{r}
sessionInfo()
```
